\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage[margin=2cm]{geometry}
\usepackage{amsfonts}
\usepackage{listings}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{color}
\usepackage{todonotes}
\lstset{
	basicstyle=\ttfamily,
	language=C++,
	backgroundcolor=\color[gray]{0.9},
	numbers=left, numberstyle=\footnotesize, stepnumber=5, 
	numbersep = 5pt, numberfirstline=false, firstnumber=last
}

\author{It prices options! \\ Luke Armitage}\date{}
\title{{\bfseries End-User Instructions} for~Two-Dimensional Black-Scholes model Option Pricer}


\begin{document}

\maketitle
This software \todo[inline]{describe what the program does.\\
brief description of methods used in software.}


\section{Running the software}
Unzip the folder \texttt{[folder name].zip} into literally wherever.
\todo[inline]{specify how to open and run software}


\section{Calibrating the model}
To calibrate the model, replace the file \texttt{data.csv}, found in the main directory, with your historical stock prices in \texttt{.csv} file format. Upon running the program, you will be prompted to specify a time interval, which is the period between recorded historical stock prices. 

The error message ``\texttt{Error opening file. Closing program.}'' means that the program could not find the file \texttt{data.csv}. Please ensure the historical stock data is in the correct file format and saved in the correct directory.



\end{document}